Blum axioms and nondeterministic computation of functions

### 2016-04-06 paper.tex
paper.tex: Add Jerusa's name as author.
paper.tex: Rewrite the Gödel numberings as a reasonableness criterion
This commit removes the mention of Gödel's incompleteness theorem,
going directly to the point of needing encodings in the Halting problem.

\author{Tiago Royer\inst{1}}

\address{
    UFSC/USP?
}

@@

\section{Quality Standards: Blum Axioms}
\label{sec:quality-standards}

@@

Sections \ref{sec:godel_numberings} and~\ref{sec:blum-axioms}
refine this concept.

\section{Gödel numberings}
\label{sec:godel_numberings}

% TODO: Note that this is a simplification

Gödel's famous Incompleteness Theorem says that
any correct and consistent formal system that includes the Peano arithmetic
will have statements that are true under that system,
but \emph{cannot be proven} under that system.
In essence,
in the proof of this theorem,
Gödel created self-referential proposition $\varphi$
that essentially says ``$\varphi$ is not provable''.
Then, if $\varphi$ is false,
by it's own assertion, it must be provable;
but, as the system is consistent,
this would mean that $\varphi$ is true
--- a contradiction.
Hence, $\varphi$ must be true.
But, as it is true,
there cannot be a proof under the current formal system to it
--- that is exactly what $\varphi$ asserts.
So,
we get a proposition we know it is true,
but we cannot prove it using only tools of that system.

We can have a glimpse at how Gödel managed to construct a formula like $\varphi$
by analyzing how Turing constructed a language
that cannot be decided by Turing machines.

The language built by Turing is known today as the halting problem.
In essence,
you are given a Turing machine $M$ and an input $w$
and must decide whether $M$ eventually halts when processing $w$.
To prove that this problem is undecidable,
Turing had first to somehow encode this problem as a language
(because all a Turing machine can understand are sentences of a language);
the key point is that the Turing machine formalism
had to be encoded into words that could be fed to Turing machines.
Thus we could pick any machine and feed its own description as an input;
this form of self-reference is analogous
to Gödel's unprovable but true sentence.

In our context,
a \emph{Gödel numbering} will simply be a means
of associating every word $w$ of $\{0, 1\}^*$
with a computable function.
Section~\ref{sec:acceptable-godel-numberings} shows a refinement of this notion,
and section~\ref{sec:blum-axioms}
defines the Blum axioms over the refined definition.

\subsection{Acceptable Gödel numberings}
\label{sec:acceptable-godel-numberings}

\begin{definition}
    Let $\mathcal P$ be the the set of partial computable functions.
    A \emph{Gödel numbering}
    is a function $\phi: \{0, 1\}^* \to \mathcal P$
    that is surjective.
\end{definition}

A Gödel numbering formalizes the concept of associating a word $w$ of $\{0, 1\}^*$
with a partial recursive function $f$ of $\mathcal P$.
We can think of $w$ as being the ``source code'' for the ``program'' $f$;
we denote $f = \phi(w)$ by $\phi_w$.

Because Gödel numberings are infinite mathematical objects,
Turing machines cannot manipulate them directly.
We thus need some technique to manipulate the numbering indirectly.

Call $\pi$ the standard Gödel numbering
that corresponds to the standard representation of Turing machines;
that is, if $w$ is the encoding of a Turing machine $M$,
then $\pi_w$ is the partial function computed by $M$.
We can manipulate $\pi$ indirectly, for instance,
by means of a universal Turing machine $U$.
Given a program $w$ and an input $x$,
if $\pi_w(x)$ exists,
then the output of the machine $U$ when its input is the pair $(w, x)$
is exactly $\pi_w(x)$.
Therefore,
even tough we did not directly manipulate the numbering $\pi$
or calculate $\pi_w$,
we did manage to indirectly compute the value of $\pi_w(x)$.
This is the \emph{universal machine theorem}.

We can also manipulate the source code directly.
Suppose $w$ represents a Turing machine that computes a function of two variables $f$.
Given two inputs $x$ and $y$, we can compute $f(x, y) = \pi_w(x, y)$
using the universal Turing machine.
If we are given a single input $x$ and leave the second variable free,
we have a function $g$ of one variable, such that
\begin{equation*}
    g(y) = f(x, y).
\end{equation*}
That is, we fixed the first input to be $x$.

This function $g$ is computable;
in fact, we can obtain a source code $w'$ for $g$ from $w$
--- just modify the machine represented by $w$
so that, on input $y$,
it writes $x$ on the tape and runs the original algorithm.%
\footnote{
    Notice we are not manipulating the machine directly,
    only its representation as a string of bits.
}
This kind of modification can be done for any source code $w$
and any input $x$;
there is a computable function $\sigma$ of two variables
that is defined for every $w$ and $x$ such that,
for every $y$,
$\pi_w(x, y)$ is defined if and only if $\pi_{\sigma(w, x)}(y)$ is defined,
and, in this case,
\begin{equation*}
    \pi_w(x, y) = \pi_{\sigma(w, x)}(y).
\end{equation*}
This is the \emph{$S_{mn}$ theorem}.%
\footnote{
    The ``full version'' of this theorem asserts that,
    for any $m$ and $n$,
    there is a computable function $\sigma$ that,
    given a program $x$ for a function of $m + n$ variables
    and $n$ chosen values $y_1, \dots, y_m$,
    \begin{equation*}
        \pi_{\sigma(x, y_1, \dots, y_m)}(z_1, \dots, z_n) =
            \pi_x(y_1, \dots, y_m, z_1, \dots, z_n);
    \end{equation*}
    hence its name \cite[p.~23]{Rogers1987}.
}

These two theorems are all that Blum requires from a Gödel numbering
to define his axiomatic notion of complexity.
We have a special name (due to Rogers) for such numbering.

\begin{definition}
    An \emph{acceptable Gödel numbering}
    is a Gödel numbering $\phi$
    that satisfies both the universal machine theorem
    and the $S_{mn}$ theorem \cite[p.~41]{Rogers1987}.
    That is, there are computable functions $f$ and $\sigma$
    that satisfies
    \begin{itemize}
        \item For every $w$ and $x$,
            $f(w, x)$ is defined if and only if $\phi_w(x)$ is defined,
            and, in this case,
            \begin{equation*}
                f(w, x) = \phi_w(x);
            \end{equation*}
        \item $\sigma$ is defined for every $w$ and $x$,
            and, for every $y$,
            $\phi_w(x, y)$ is defined if and only if $\phi_{\sigma(w, x)}(y)$ is defined,
            and, in this case,
            \begin{equation*}
                \phi_w(x, y) = \phi_{\sigma(w, x)}(y).
            \end{equation*}
    \end{itemize}
\end{definition}

@@

This section surveys four approaches
for defining nondeterministic computation of functions.
Sections \ref{sec:acceptable-godel-numberings} and~\ref{sec:blum-axioms}
can be understood as ``quality requirements''
for the definitions;
that is, the definitions will need to be translatable as
acceptable Gödel numberings,
and the ``natural'' extension of space and time complexity
must satisfy Blum axioms.

>>>>

\author{Tiago Royer\inst{1}, Jerusa Marchi\inst{1}}

\address{
    Universidade Federal de Santa Catarina --- Departamento de Informática e Estatística
}

@@

\section{Quality Standards}
\label{sec:quality-standards}

@@

Section~\ref{sec:godel-numberings}
presents a formalization of the ``computation'' part
and section~\ref{sec:blum-axioms}
presents a formalization of the ``non-deterministic'' part.

\subsection{Reasonableness criterion: acceptable Gödel numberings}
\label{sec:godel-numberings}

One of the most important theoretical results concerning Turing machines
is the existence of undecidable problems.
Namely, the \emph{halting problem}
(the task of deciding whether a given Turing machine will halt on a given input)
cannot be solved by Turing machines.
To formalize (and prove) this fact,
we must first decide how to encode Turing machines in strings
(because all a Turing machine can manipulate are strings).

Each Turing machine can be associated
to the corresponding partial recursive function it computes.
There are several ways to encode Turing machines as strings,
but what is most important about them is that they allow us
to manipulate these partial recursive functions \emph{indirectly}
--- partial recursive functions are (potentially) infinite objects,
so we cannot write them down on a Turing machine tape,
but we \emph{can} write the encoding of a Turing machine that compute these functions.

Therefore,
these encodings provide a way to associate a string
(which is a finite, manipulable object)
with a partial recursive function
(which is an infinite, mathematical, ``untouchable'' object).
An encoding is an \emph{enumeration} of all partial recursive function.

In a way,
we are going the opposite direction:
instead of associating functions with strings,
we want to associate machines with functions.
We will use the concept of \emph{acceptable Gödel numbering}
(\cite[p.~41]{Rogers1987},~\cite[p.~324]{Blum1967})
to impose reasonableness criteria
to our definitions.

\begin{definition}
    Let $\mathcal P$ be the set of all partial recursive functions.
    An \emph{acceptable Gödel numbering}
    is a function $\phi : \{0, 1\}^* \to \mathcal P$,
    that associates each string (or program) $w \in \{0, 1\}^*$
    to a function $\phi_w \in \mathcal P$,
    that satisfies
    \begin{enumerate}
        \item $\phi$ is surjective;
            that is, every partial recursive function $f \in \mathcal P$
            has a program $w \in \{0, 1\}^*$ such that $\phi_w = f$;
            \label{surjectiveness}
        \item There is a partial recursive function
            $U:\{0, 1\}^* \times \{0, 1\}^* \to \{0, 1\}^*$ such that,
            for every $w$ and $x$,
            $U(w, x)$ is defined if and only if $\phi_w(x)$ is defined,
            and, in this case,
            \begin{equation*}
                U(w, x) = \phi_w(x);
            \end{equation*}
            \label{universal-tm}
        \item There is a total recursive function
            $\sigma:\{0, 1\}^* \times \{0, 1\}^* \to \{0, 1\}^*$ such that,
            for every $y$,
            $\phi_w(x, y)$ is defined if and only if $\phi_{\sigma(w, x)}(y)$ is defined,
            and, in this case,
            \begin{equation*}
                \phi_w(x, y) = \phi_{\sigma(w, x)}(y).
            \end{equation*}
            \label{smn-theorem}
    \end{enumerate}
\end{definition}

This concept captures the intuitive notion of
``well-behaved numbering''.

Condition~\ref{surjectiveness}
guarantees that the image of~$\phi$ is all of~$\mathcal P$,
so that the numbering neither ``forgets'' a partial recursive function
nor generate a function that is not partial recursive.

Condition~\ref{universal-tm} is the universal Turing machine theorem.
By varying the first input of~$U$,
we can obtain any partial recursive function in a controlled manner.

And Condition~\ref{smn-theorem} is known as the $S_{mn}$ theorem,
or the iteration theorem.

@@

This section surveys four approaches
for defining nondeterministic computation of functions.
Sections \ref{sec:godel-numberings} and~\ref{sec:blum-axioms}
can be understood as ``quality requirements''
for the definitions;
that is, the definitions will need to be translatable as
acceptable Gödel numberings,
and the ``natural'' extension of space and time complexity
must satisfy Blum axioms.
